What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2022-12-28 paper/main.tex
added cls and bst files

@@

    \author[inst1]{Runming Li}
    \ead{runmingl@andrew.cmu.edu}

    \author[inst1]{Keerthana Gurushankar}
    \ead{kgurusha@andrew.cmu.edu}

@@

    \author[inst1]{Marijn Heule\corref{cor1}}
    \ead{marijn@cmu.edu}

@@

    \cortext[cor1]{Corresponding author}

@@

    \begin{highlights} %TODO
        \item Research highlight 1
        \item Research highlight 2
    \end{highlights}

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_{sf}$), and for those for paths for looping from the final state back to itself (say $r_{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_{sf}r_{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in digraph of the automaton.

@@

\begin{algorithm}[h!]
    \caption{reduce\_dfa}
    \begin{algorithmic}
        \Require ($G, v$), a DFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_{loop}}v$}
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in} r_{loop}* r_{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in}  r_{out}} w$
            \EndIf
        \EndFor
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_dfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_{ss}| r_{sf} r_{ff}^* r_{fs})^* r_{sf} r_{ff}$
        % if called by first visit, this is r_{ss}* r_{sf}
        % if called by ba2wregex, s=f, this is r_{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_dfa does its input to output behaviour
    For any DFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_dfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_dfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{$u$ to $w$ path in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_{uv}}v\xrightarrow{r_{vw}}w$, with the trace $r=r_{uv}r_{vw}$. If there is a loop edge $r_{vv}$, this edge can be travelled any number of times before exiting $v$, thus the regular expression is $r=r_{uv}r_{vv}^*r_{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

\begin{lemma}
    For any DFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{dfa2regex}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$ (i.e., the language accepted by $G$)
\end{lemma}
\begin{proof}
    The while loop clearly terminates since the number of interior nodes is strictly decreasing, and upon termination, $G$ is a DFA containing only the nodes $s$ and $f$. Now, every non-terminal visit from $s$ to $f$ and back to $s$, can be replaced by an edge $s\xrightarrow{r_{sf}r_{ff}^* r_{fs}} s$, instead of the edge $f\to s$. This DFA yields the regular language $(r_{ss}|r_{sf}r_{ff}^* r_{fs})^* r_{sf} r_{ff}$, as outputted by the algorithm.
\end{proof}

@@

\section{Tool showcase}
To demonstrate the result of our tool, we present two examples, one from a real-world model checking example in \cite{ZR14} and one from a randomly generated \ltl\ formula.

\begin{example}
    \cite{ZR14} presents a model verification specification says ``[i]f a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft'' which corresponds to \ltl\ formula
    \[
        \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
    \]
    For simplicity, we swap the concrete atomic proposition to $a, b, c$ and get
    \[
        \always (a \land b \limplies \nextt (\neg b))
    \]
    For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex14}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex14/ex14.png}
        \caption{Timeline for $\always (a \land b \limplies \nextt (\neg b))$}
        \label{fig:ex14}
    \end{figure}
\end{example}

>>>>

\usepackage{graphicx}

@@

	\author[inst1]{Runming Li\corref{cor1}}
    \ead{runmingl@andrew.cmu.edu}

	\author[inst1]{Keerthana Gurushankar\corref{cor1}}
    \ead{kgurusha@andrew.cmu.edu}

@@

    \author[inst1]{Marijn Heule\corref{cor2}}
    \ead{marijn@cmu.edu}

@@

	\cortext[cor1]{These authors contributed equally to this work}
    \cortext[cor2]{Corresponding author}

@@

    \begin{highlights} %TODO ~ Highly encouraged but not mandatory
        \item Research highlight 1
        \item Research highlight 2
    \end{highlights}

@@

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

@@

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_\mathrm{sf}$), and for those for paths for looping from the final state back to itself (say $r_\mathrm{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_\mathrm{sf}r_\mathrm{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in digraph of the automaton.

@@

\begin{algorithm}[h!]
    \caption{reduce\_dfa}
    \begin{algorithmic}
        \Require ($G, v$), a DFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_\mathrm{loop}}v$}
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in} r_\mathrm{loop}* r_\mathrm{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_\mathrm{in}} v, v \xrightarrow{r_\mathrm{out}} w$ with $u \xrightarrow{r_\mathrm{in}  r_\mathrm{out}} w$
            \EndIf
        \EndFor
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_dfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_\mathrm{ss}| r_\mathrm{sf} r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$
        % if called by first visit, this is r_\mathrm{ss}* r_\mathrm{sf}
        % if called by ba2wregex, s=f, this is r_\mathrm{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

@@

\section{Theoretical Analyses}
\subsection{Correctness of Regex Translation}
\begin{lemma}
    % reduce_dfa does its input to output behaviour
    For any DFA $G$ and any state $v$ in $G$ that is neither a start state or a final state, \textbf{reduce\_dfa}$(G, v)$ preserves the regular language accepted by $G$
\end{lemma}
\begin{proof}
    Let $G'$ be the graph of $G$ post reduction by the application of \textbf{reduce\_dfa}$(G, v)$. We show that the trace of every path accepted by $G$ is also accepted by $G'$. Suppose a path accepted by $G$ does not pass through $v$, clearly this is true. Else, suppose it passes through $v$, since $v$ is an interior node, $v$ cannot be the first or last in the path. Thus, for every pass through $v$, let $u\neq v$ be the last node passed before entering $v$, and likewise $w$ be the first node after exiting $v$. We show that the regular language of sub-traces from $u$ to $w$ in $G$ (shown in \cref{fig:uvw-dfa}) is identical to that in $G'$.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.75]{img/uvw_dfa.pdf}
        \caption{$u$ to $w$ path in $G$}
        \label{fig:uvw-dfa}
    \end{figure}
    Suppose there is no loop at $v$, the path from $u$ to $w$ must simply be $u\xrightarrow{r_\mathrm{uv}}v\xrightarrow{r_\mathrm{vw}}w$, with the trace $r=r_\mathrm{uv}r_\mathrm{vw}$. If there is a loop edge $r_\mathrm{vv}$, this edge can be travelled any number of times before exiting $v$, thus the regular expression is $r=r_\mathrm{uv}r_\mathrm{vv}^*r_\mathrm{vw}$. This edge $u\xrightarrow{r}w$ is added in $G'$. Thus, for every path accepted by $G$, for every subpath entering and exiting $v$ from some $u$ to $w$, there exists a corresponding subpath from $u$ to $w$ in $G'$, and thus a corresponding path accepted by $G'$.
\end{proof}

\begin{lemma}
    For any DFA $G$ with start state $s$ and (exactly one) final state $f$, \textbf{dfa2regex}$(G, s, f)$ outputs the regular expression of all paths from $s$ to $f$ (i.e., the language accepted by $G$)
\end{lemma}
\begin{proof}
    The while loop clearly terminates since the number of interior nodes is strictly decreasing, and upon termination, $G$ is a DFA containing only the nodes $s$ and $f$. Now, every non-terminal visit from $s$ to $f$ and back to $s$, can be replaced by an edge $s\xrightarrow{r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs}} s$, instead of the edge $f\to s$. This DFA yields the regular language $(r_\mathrm{ss}|r_\mathrm{sf}r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$, as outputted by the algorithm.
\end{proof}

@@

\section{Tool showcase}
To demonstrate the result of our tool, we present two examples, one from a real-world model checking example from Zhao and Rozier~\cite{ZR14} and one from a randomly generated \ltl\ formula.

\begin{example}
    Zhao and Rozier~\cite{ZR14} presents a model verification specification says ``[i]f a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft'' which corresponds to \ltl\ formula
    \[
        \always (\texttt{tsafe.TSAFE\_command1} \land \texttt{controller.CTR\_control\_1} \limplies \nextt (\neg \texttt{controller.CTR\_control\_1}))
    \]
    For simplicity, we swap the concrete atomic proposition to $a, b, c$ and get
    \[
        \always (a \land b \limplies \nextt (\neg b))
    \]
    For this \ltl\ formula, our tool generates the timeline representation in \cref{fig:ex14}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex14/ex14.png}
        \caption{Timeline for $\always (a \land b \limplies \nextt (\neg b))$}
        \label{fig:ex14}
    \end{figure}
\end{example}
